---
title: "GALD liver cell line"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
library(tidyverse)
library(limma)
library(broom)
library(ggrepel)
library(preprocessCore)
library(ggpubr)
library(openxlsx)
```

Differential abundance analysis in MS data from immunoprecipitations from IPs with healthy and GALD plasma. 
Plasma from a patient with primary biliary cirrhosis (PBC) serves as positive control. 

```{r}
proteinGroups <- read_tsv("../data/20210104/proteinGroups.txt")
proteinGroups
```

Design matrix

```{r}
expOverview <- readxl::read_excel("../data/experimental_overview_manuscript.xlsx", sheet = "experimental_design_Huh-7")
design <- expOverview[, c("sample", "condition")]
design$condition <- factor(design$condition, levels = c("GALD", "healthy", "PBC"))
design
```

## QC

Filter out contaminants and matches to the decoy database.  
Remove protein groups that were identified with no unique peptide. 

```{r}
proteinGroups <- proteinGroups %>%
  filter(is.na(`Potential contaminant`),
         is.na(Reverse),
         `Unique peptides` != 0)
```

## Quantile Normalization

Rename labels into sample names and count identified proteins.

```{r}
# replace labels in column names with sample names
proteinColnames <- names(proteinGroups)
for (i in seq(dim(expOverview)[1])) {
  proteinColnames <- gsub(
    paste0(" ", expOverview$label[i], " "),
    paste0(" ", expOverview$sample[i], " "),
    proteinColnames
  )
}

colnames(proteinGroups) <- proteinColnames

intensColNames <- colnames(proteinGroups)[startsWith(colnames(proteinGroups), "Reporter intensity corrected")]

numQuantifiedProteins <- proteinGroups %>%
  rowwise() %>%
  mutate(sum_intensities = sum(!!!syms(intensColNames))) %>%
  filter(sum_intensities != 0) %>%
  dim()

cat("# identified protein groups", dim(proteinGroups)[1], "\n")
cat("# quantified protein groups", numQuantifiedProteins[1])
```


## Data transformation and normalization

Log2 transform intensities.

```{r}
intensLog2ColNames <- paste0(intensColNames, "_log2")

proteinGroups <- proteinGroups %>%
  mutate(across(c(!!!syms(intensColNames)), .fns = list(log2 = ~(log2(.))))) %>%
  # replace -Inf with NA
  mutate_at(vars(intensLog2ColNames), list(~na_if(., -Inf)))
```


Plot corrected intensities before normalization.

```{r}
plot_intensities <- function(proteinGroups, colNames) {
  proteinGroups %>%
    select(`Protein IDs`, colNames) %>%
    pivot_longer(cols = -`Protein IDs`) %>%
    rowwise() %>%
    mutate(sample = strsplit(name, split = " ")[[1]][4]) %>%
    ggplot(aes(x = sample, y = value)) +
    geom_violin() +
    theme_bw() +
    theme(axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    ))
}

plot_intensities(proteinGroups, intensLog2ColNames) +
  ylab("log2 Reporter intensity corrected")
```

Apply quantile normalization.

```{r}
# perform quantile normalization
proteinGroupsIntensLog2Norm <- as.data.frame(normalize.quantiles(as.matrix(select(proteinGroups, intensLog2ColNames))))
# add column and row names back
proteinGroupsIntensLog2Norm <- as_tibble(add_column(proteinGroupsIntensLog2Norm, 
                                                    Protein_IDs=proteinGroups$`Protein IDs`, 
                                                    .before="V1"))
intensLog2NormColNames <- paste0(intensLog2ColNames, "_norm")
colnames(proteinGroupsIntensLog2Norm) <- c("Protein IDs", intensLog2NormColNames)

# combine with other table
proteinGroups <- merge(proteinGroups, proteinGroupsIntensLog2Norm, by="Protein IDs")
```

Plot after quantile normalization.

```{r}
plot_intensities(proteinGroups, intensLog2NormColNames) +
  ylab("log2 normalized Reporter intensity corrected")
```

## PCA

PCA to check if samples can be separated into diseased and controls.

```{r}
# color palette from Won https://davidmathlogic.com/colorblind/#%23000000-%23E69F00-%2356B4E9-%23009E73-%23F0E442-%230072B2-%23D55E00-%23CC79A7
colorPalette <- c("#E69F00", "#009E73", "#0072B2")
colorPaletteVolcano <- c("#56B4E9", "#D55E00")

plot_pca <- function(proteinGroups, colNames) {
  pca.log2 <- proteinGroups %>%
    select(colNames) %>%
    drop_na() %>%
    t() %>%
    prcomp()
  
  percent_variance_log2 <-
    summary(pca.log2)$importance["Proportion of Variance", ] * 100
  
  plot.log2 <- as_tibble(pca.log2$x) %>%
    mutate(Sample = rownames(pca.log2$x),
           Condition = design$condition)
  
  ggplot(plot.log2, aes(x = PC1,
                        y = PC2,
                        color = Condition)) +
    geom_point() +
    scale_color_manual(values = colorPalette) +
    xlab(label = paste0("PC1 (", round(percent_variance_log2[1], 2), "% variance)")) +
    ylab(label = paste0("PC2 (", round(percent_variance_log2[2], 2), "% variance)")) +
    theme_bw()
}

plot_pca(proteinGroups, intensLog2ColNames)
```


```{r}
g <- plot_pca(proteinGroups, intensLog2NormColNames)

# ggsave(g, filename = "../plots/cell_line_pca_norm.pdf", width = 4, height = 2.5)
# ggsave(g, filename = "../plots/cell_line_pca_norm.png", width = 4, height = 2.5)
g
```

After quantile normalization, PCA separates GALD samples, negative controls and the positive control on PC1 and PC2. 

```{r}
plot_proteins <- function(selected_Protein_IDs, unique_peptides=T) {

  proteinGroupsPlotting <- proteinGroupsIntensLog2Norm %>%
    merge(proteinGroups[, c("Protein IDs", "Gene names", "Unique peptides")], by = "Protein IDs", all.x = T) %>%
    pivot_longer(cols = intensLog2NormColNames, values_to = "log2_intens") %>%
    rowwise() %>%
    mutate(sample = strsplit(name, split=" ")[[1]][4]) %>%
    merge(design, by="sample") %>%
    mutate(gene_info = paste0(`Gene names`, " (", `Unique peptides`, ")"))

  plot_data <- proteinGroupsPlotting %>%
    filter(`Protein IDs` %in% selected_Protein_IDs) %>%
    group_by(`Protein IDs`) %>%
    mutate(log2_intens_aligned = log2_intens - median(log2_intens)) %>%
    arrange(match(`Protein IDs`, selected_Protein_IDs))
  
  if (unique_peptides) {
    gene_levels <- unique(plot_data$gene_info)
    g <- proteinGroupsPlotting %>%
      filter(`Protein IDs` %in% selected_Protein_IDs) %>%
      mutate(gene_info = factor(gene_info, levels = gene_levels)) %>%
      ggplot(aes(x=condition, y=log2_intens, color=condition)) +
      geom_point(alpha=0.5) +
      facet_wrap(~gene_info, scales="free_y")
  } else {
    gene_levels <- unique(plot_data$`Gene names`)
    g <- proteinGroupsPlotting %>%
      filter(`Protein IDs` %in% selected_Protein_IDs) %>%
      mutate(`Gene names` = factor(`Gene names`, levels = gene_levels)) %>%
      ggplot(aes(x=condition, y=log2_intens, color=condition)) +
      geom_point(alpha=0.5) +
      facet_wrap(~`Gene names`, scales="free_y")
  }
  
  g <- g +
    scale_color_manual(values=colorPalette) +
    theme_bw() +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("normalized log2 intensity")

  return(g)
}
```

## Limma analysis for differentially abundant protein groups

```{r}
# prepare input for limma
# clean column names
proteinColnames <- names(proteinGroupsIntensLog2Norm)
for (i in seq(dim(expOverview)[1])) {
  proteinColnames <- gsub(
    paste0(".*", expOverview$sample[i], ".*"),
    paste0(expOverview$sample[i]),
    proteinColnames
  )
}

limmaInput <- proteinGroupsIntensLog2Norm
colnames(limmaInput) <- proteinColnames

# exclude positive control from analysis
limmaInput <- limmaInput %>%
  select(-PBC) %>%
  as.data.frame()

# turn column with protein IDs into rownames
rownames(limmaInput) <- limmaInput$`Protein IDs`
limmaInput <- limmaInput[2:8]

# design matrix and contrast
limmaDesign <- as.character(pull(design[design$sample != "PBC",], condition))
limmaDesign <- model.matrix( ~ 0 + limmaDesign)
limmaContrasts <- makeContrasts(limmaDesignGALD - limmaDesignhealthy,
                                levels = colnames(limmaDesign))

# fit limma model
fit <- lmFit(limmaInput, limmaDesign)
fit <- contrasts.fit(fit, limmaContrasts)
fit <- eBayes(fit)
```

Merge limma results with protein groups table from MaxQuant. 

```{r}
limmaResult <- topTable(fit, n = dim(proteinGroupsIntensLog2Norm)[1])
limmaResult$`Protein IDs` <- rownames(limmaResult)
# merge gene names
limmaResult <- merge(limmaResult,
                     proteinGroups,
                     by = "Protein IDs",
                     all.x = T)
```

### Limma p-value distribution

Usually expected to be uniform but since this is from IPs with fewer proteins it is not the case here. 

```{r}
limmaResult %>%
  ggplot(aes(P.Value)) +
  geom_histogram(binwidth = 0.05, color = "black") +
  ggtitle("P-value distribution") +
  theme_bw()
```

### Limma Volcano plot

Volcano plot to visualize significant hits. 

```{r}
gVolcano <- ggplot(limmaResult,
                   aes(
                     x = logFC,
                     y = -log10(P.Value),
                     color = adj.P.Val < 0.05
                   )) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = colorPaletteVolcano) +
  geom_vline(xintercept = 0,
             alpha = 0.75,
             linetype = "dashed") +
  geom_text_repel(data = filter(limmaResult, adj.P.Val < 0.05),
                  aes(label = `Gene names`)) +
  theme_bw() +
  theme(legend.position = "none")

# ggsave(gVolcano, filename = "../plots/cell_line_limma_volcano.pdf", width = 6, height = 4)
# ggsave(gVolcano, filename = "../plots/cell_line_limma_volcano.png", width = 6, height = 4)
gVolcano
```

### Limma best hits

Look at proteins that are significantly differentially abundant. 

```{r}
topHitsProteinIds <- limmaResult %>%
  filter(adj.P.Val < 0.05) %>%
  arrange(logFC)

topHitsProteinIds
  
gTopHits <- plot_proteins(pull(topHitsProteinIds, `Protein IDs`))

# ggsave(gTopHits, filename = "../plots/cell_line_sign_hits_limma_indiv.pdf", width = 6, height=4)
# ggsave(gTopHits, filename = "../plots/cell_line_sign_hits_limma_indiv.png", width = 6, height=4)
gTopHits

proteinGroups %>%
  filter(`Protein IDs` %in% pull(topHitsProteinIds, `Protein IDs`)) %>%
  select(`Protein IDs`, `Gene names`, `Unique peptides`)

topHitsLimmaTable <- topHitsProteinIds %>%
  select(`Majority protein IDs`, `Gene names`, `Protein names`,  P.Value, adj.P.Val, logFC, `Unique peptides`) %>%
  arrange(logFC)

topHitsLimmaTable
```

```{r}
g <- ggarrange(gVolcano, gTopHits, labels = c("A", "B"))
# ggsave(g, filename = "../plots/cell_line_limma_results.png", width = 8, height = 3.5)
# ggsave(g, filename = "../plots/cell_line_limma_results.pdf", width = 8, height = 3.5)
g
```

## Positive PBC control

List of PBC antigens from manuscript.

Only 5 out of 6: P09622 (Gene name DLD) Dihydrolipoyl dehydrogenase, mitochondrial not detected. 

```{r}
PbcAntigens <- proteinGroups %>%
  filter(grepl("P10515|P09622|P11177|P08559|O00330|O60232", `Protein IDs`)) %>%
  pull(`Protein IDs`)

proteinGroups %>%
  filter(grepl("P09622", `Protein IDs`)) %>%
  pull(`Protein IDs`)

proteinGroups %>%
  filter(grepl("DLD", `Gene names`)) %>%
  pull(`Protein IDs`)

PbcAntigens
```

Plot intensities of the antigens. 

```{r}
g <- plot_proteins(PbcAntigens, unique_peptides = F)

# ggsave(g, filename = "../plots/cell_line_PBC_antigens_indiv.pdf", width = 6, height=4)
# ggsave(g, filename = "../plots/cell_line_PBC_antigens_indiv.png", width = 6, height=4)
g
```

Calculate fold change between PBC patient and average of controls. 

```{r}
healthyCols <- c("Reporter intensity corrected healthy1 170912_8plex_scx_log2_norm", 
                 "Reporter intensity corrected healthy2 170912_8plex_scx_log2_norm", 
                 "Reporter intensity corrected healthy3 170912_8plex_scx_log2_norm")

proteinGroupsIntensLog2Norm %>%
  filter(`Protein IDs` %in% PbcAntigens) %>%
  rowwise() %>%
  mutate(ctrl_mean = mean(!!!syms(healthyCols)),
         log2FC = `Reporter intensity corrected PBC 170912_8plex_scx_log2_norm` - ctrl_mean)
```

Save the full results table.

```{r}
suplTableManuscript <- merge(proteinGroups,
                             limmaResult[, c("Protein IDs",
                                             "logFC",
                                             "AveExpr",
                                             "t",
                                             "P.Value",
                                             "adj.P.Val",
                                             "B")],
                             by = "Protein IDs")
# write.xlsx(suplTableManuscript, file = "../outputs/Huh-7_MS_results.xlsx")
```


```{r}
sessionInfo()
```
